---
title: "Coherent multi-disease, multi-stratum mortality projection (Japan)"
author: "bamp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coherent multi-disease, multi-stratum mortality projection (Japan)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.4, dpi = 96)
options(mc.cores = 1L)
set.seed(1)
```

```{r lib}
library(bamp)
```

## The problem

Projecting mortality for several diseases, by sex, the *independent* way — one model per
disease, one model per sex — has two well-known failures:

1. **Incoherence across diseases.** Independently projected cause-specific rates do not add up
   to a sensible all-cause trajectory, and the dependence between causes (cardiovascular causes
   tend to rise and fall *together*) is ignored.
2. **Divergence across strata.** Independent sex-specific forecasts drift apart without bound,
   producing implausible male/female gaps at long horizons.

This vignette uses Japanese mortality (1995–2019, ages 30–80+) for **CHD**, **stroke** and
**all other ("non-modelled") causes** — three competing causes that partition all deaths
(`non-modelled = total − CHD − stroke`) — to show the **recommended** coherent approach and
contrast it with independent projections. We project to **2050** using the official population
projections as future denominators.

```{r data}
## ---- load the bundled Japan data, ages 30-80+, [year x agegroup] ----
jdir <- system.file("extdata", "japan", package = "bamp")
if (jdir == "" || !dir.exists(jdir)) jdir <- "../inst/extdata/japan"   # source tree (build time)
age10 <- c("age30_34","age35_39","age40_44","age45_49","age50_54","age55_59",
           "age60_64","age65_69","age70_74","age75_79")
age80p <- c("age80_84","age85_89","age90_94","age95_99","age100_")
read_death <- function(f) {
  d <- read.csv(file.path(jdir, f), check.names = FALSE)
  M <- cbind(as.matrix(d[, age10]), "80+" = rowSums(d[, age80p]))      # collapse 80+ to match pop
  rownames(M) <- d[["year"]]; M
}
pop_mat <- function(sx) {
  p <- read.csv(file.path(jdir, "population_per_5yo.csv"))
  m <- xtabs(pops ~ Year + age, data = p[p$sex == sx & p$age %in% seq(30, 80, 5), ])
  matrix(as.double(m), nrow(m), ncol(m), dimnames = dimnames(m))
}
load_sex <- function(sx, fit = 1995:2019, to = 2050) {
  yr <- as.character(fit)
  chd <- read_death(sprintf("chd_%s.csv", sx))[yr, ]
  str <- read_death(sprintf("stroke_%s.csv", sx))[yr, ]
  oth <- read_death(sprintf("nonmodeled_%s.csv", sx))[yr, ]
  pop <- pop_mat(sx)
  list(CHD = chd, stroke = str, other = oth, total = chd + str + oth,
       pop = pop[yr, ], pop_fut = pop[as.character(min(fit):to), ])
}
men   <- load_sex("men")
women <- load_sex("women")
ages  <- c("30-34","35-39","40-44","45-49","50-54","55-59","60-64","65-69","70-74","75-79","80+")
years <- 1995:2050; A <- 9   # focus age band 70-74
cat(sprintf("grid: %d age groups x %d fitted years (1995-2019), projected to 2050\n",
            length(ages), nrow(men$total)))
```

```{r plot-helpers, include = FALSE}
band <- function(x, q, col, add = FALSE, lty = 1, ...) {   # q: [quantile, period] (lo, mid, hi)
  if (!add) plot(x, q[2, ], type = "n", ...)
  polygon(c(x, rev(x)), c(q[1, ], rev(q[3, ])), col = adjustcolor(col, 0.18), border = NA)
  lines(x, q[2, ], col = col, lwd = 2.2, lty = lty)
}
mco <- list(iterations = 1500, burn_in = 500, thin = 2)            # prototype MCMC (modest, for a vignette)
bo  <- list(number_of_iterations = 3000, burn_in = 1000, step = 2, tuning = 300)
```

## Part A — competing diseases, coherently (men)

The recommended model for several competing causes is `bamp_multicause()`. It stick-breaks the
deaths into cause-specific shares, so the cause rates **sum to all-cause by construction**, and it
couples the cause *trends* through a cross-cause covariance, so it learns whether causes move
together. Cause-specific **hazards** (additive, per person-year) are what a competing-risk
life-table / microsimulation needs.

```{r fit-multicause}
fit_mc <- bamp_multicause(list(CHD = men$CHD, stroke = men$stroke, other = men$other),
                          men$pop, age = "rw1", period = "rw1", cohort = "rw1",
                          periods_per_agegroup = 5, order = 1:3, mcmc = mco)
pmc <- predict_multicause(fit_mc, periods = 31, population = men$pop_fut,
                          hazard = TRUE, period_length = 1)
```

### Cause-specific projections (and they stay coherent)

```{r plot-causes, fig.height = 4.6}
cols <- c(CHD = "#c0392b", stroke = "#2980b9", other = "#27ae60")
yl <- range(unlist(lapply(names(cols), function(nm) pmc[[nm]]$rate[c(1, 3), , A])))
plot(NA, xlim = c(1995, 2050), ylim = yl, log = "y", xlab = "year",
     ylab = "mortality rate, log scale (age 70-74)",
     main = "Men: cause-specific mortality, 1995-2050")
abline(v = 2019, lty = 3, col = "grey50")
for (nm in names(cols)) {
  band(years, pmc[[nm]]$rate[, , A], cols[nm], add = TRUE)
  points(1995:2019, men[[nm]][, A] / men$pop[, A], pch = 16, cex = 0.5, col = cols[nm])
}
legend("bottomleft", names(cols), col = cols, lwd = 2.2, bty = "n")
text(2019, yl[2], "fit | project", pos = 2, cex = 0.8, col = "grey40")
```

CHD and stroke both decline; "other" causes dominate and flatten. Dots are the observed rates.
Because the model is coherent, the three cause rates **add up to the all-cause rate every draw**:

```{r coherence-check}
cat("max |sum of cause rates - all-cause| over the whole grid:", signif(pmc$coherence_maxerr, 3), "\n")
```

### What independent cause models miss #1: coherence with all-cause

If instead we fit each cause *independently* and add the projected rates, the sum drifts away
from a direct all-cause projection — there is no guarantee the parts add to the whole.

```{r incoherence, fig.height = 4.4}
fit_ind <- lapply(c("CHD", "stroke", "other"), function(nm)
  bamp(men[[nm]], men$pop, age = "rw1", period = "rw1", cohort = "rw1",
       periods_per_agegroup = 5, mcmc.options = bo, parallel = FALSE, verbose = FALSE))
pr_ind <- lapply(fit_ind, predict_apc, periods = 31, population = men$pop_fut)
sum_ind <- Reduce(`+`, lapply(pr_ind, function(p) p$pr[2, , A]))      # sum of independent medians
futA <- years >= 2019
plot(years, pmc$total$rate[2, , A], type = "l", lwd = 2.4, col = "black",
     xlim = c(2019, 2050), ylim = range(c(pmc$total$rate[2, futA, A], sum_ind[futA])),
     xlab = "year", ylab = "all-cause rate (age 70-74)",
     main = "Men, all-cause: coherent total vs sum of independent cause models")
lines(years, sum_ind, lwd = 2.4, col = "#e67e22", lty = 2)
abline(v = 2019, lty = 3, col = "grey50")
legend("topright", c("coherent (bamp_multicause)", "sum of independent cause fits"),
       col = c("black", "#e67e22"), lwd = 2.4, lty = c(1, 2), bty = "n")
```

The orange "sum of independent fits" peels away from the coherent all-cause projection — the
incoherence grows with horizon. The coherent model cannot do this: its causes are shares of the
all-cause total.

### What independent cause models miss #2: cross-cause dependence

The coupled model estimates how the cause trends move together (the cross-cause correlation of
the period innovations). Independent models implicitly assume this is **zero**.

```{r cor-omega}
round(pmc$cor_omega, 2)
```

CHD and stroke are strongly **positively** correlated — in Japan both cardiovascular causes
declined together (shared risk-factor improvements). A coupled model propagates that joint
movement into the forecast; independent models treat the causes as unrelated, mis-stating the
joint uncertainty (e.g. of total cardiovascular mortality).

## Part B — coherence across sex

For strata (here sex) the recommended model is `bamp_coherent()`: the sexes share the period and
cohort trends and differ by a *mean-reverting* deviation, so the male/female gap cannot diverge.
We compare it with two independent `bamp` fits on total (30+) mortality.

```{r fit-sex}
fc <- bamp_coherent(list(men = men$total, women = women$total),
                    list(men = men$pop,  women = women$pop),
                    age = "rw1", period = "rw1", cohort = "rw1",
                    periods_per_agegroup = 5, deviation = "ar1", mcmc = mco)
pc <- predict_coherent(fc, periods = 31, population = list(men = men$pop_fut, women = women$pop_fut))

fm <- bamp(men$total,   men$pop,   age = "rw1", period = "rw1", cohort = "rw1",
           periods_per_agegroup = 5, mcmc.options = bo, parallel = FALSE, verbose = FALSE)
fw <- bamp(women$total, women$pop, age = "rw1", period = "rw1", cohort = "rw1",
           periods_per_agegroup = 5, mcmc.options = bo, parallel = FALSE, verbose = FALSE)
pm <- predict_apc(fm, periods = 31, population = men$pop_fut)
pw <- predict_apc(fw, periods = 31, population = women$pop_fut)
```

### Projected male/female mortality

```{r plot-sex-rates, fig.height = 4.6}
par(mfrow = c(1, 2), mar = c(4, 4, 2.4, 1))
yl <- c(0, 0.05)
plot(NA, xlim = c(1995, 2050), ylim = yl, xlab = "year", ylab = "total mortality (70-74)",
     main = "Coherent (bamp_coherent)")
band(years, pc$men$rate[, , A],   "#c0392b", add = TRUE)
band(years, pc$women$rate[, , A], "#8e44ad", add = TRUE)
abline(v = 2019, lty = 3, col = "grey50"); legend("topright", c("men","women"),
       col = c("#c0392b","#8e44ad"), lwd = 2.2, bty = "n")
plot(NA, xlim = c(1995, 2050), ylim = yl, xlab = "year", ylab = "",
     main = "Independent (two bamp fits)")
band(years, pm$pr[, , A], "#c0392b", add = TRUE, lty = 2)
band(years, pw$pr[, , A], "#8e44ad", add = TRUE, lty = 2)
abline(v = 2019, lty = 3, col = "grey50")
par(mfrow = c(1, 1))
```

### The headline: the sex gap diverges independently, but not coherently

The right way to see the difference is the **uncertainty of the male/female gap** as the horizon
grows. The coherent gap is *stationary* (it plateaus); the independent gap is a random walk (it
grows without bound and overtakes the coherent one).

```{r plot-gap-sd, fig.height = 4.4}
gap_sd <- function(rm, rw) vapply(seq_along(years),
  function(t) sd(log(rm[t, A, ]) - log(rw[t, A, ])), 0.0)
sd_coh <- gap_sd(pc$men$samples$rate, pc$women$samples$rate)
sd_ind <- gap_sd(pm$samples$pr,        pw$samples$pr)
fut <- years >= 2019
plot(years[fut], sd_coh[fut], type = "l", lwd = 2.6, col = "#16a085", ylim = c(0, max(sd_ind[fut])),
     xlab = "year", ylab = "SD of male-female log-gap (age 70-74)",
     main = "Sex-gap uncertainty: coherent plateaus, independent diverges")
lines(years[fut], sd_ind[fut], lwd = 2.6, col = "#e67e22", lty = 2)
legend("topleft", c("coherent (mean-reverting)", "independent (random walk)"),
       col = c("#16a085", "#e67e22"), lwd = 2.6, lty = c(1, 2), bty = "n")
```

The independent gap (orange) keeps fanning out and crosses above the coherent gap (teal) — by
mid-century its implied male/female differences are arbitrarily wide. The coherent model holds the
gap to a plausible, stationary band. The point forecast tells the same story: the independent gap
*drifts* (it extrapolates each sex's recent slope), while the coherent gap stays put.

```{r mean-drift}
gap_mean <- function(rm, rw, t) mean(log(rm[t, A, ]) - log(rw[t, A, ]))
cat(sprintf("mean male-female log-gap (70-74)  2020 -> 2050:  coherent %.2f -> %.2f | independent %.2f -> %.2f\n",
            gap_mean(pc$men$samples$rate, pc$women$samples$rate, 26),
            gap_mean(pc$men$samples$rate, pc$women$samples$rate, 56),
            gap_mean(pm$samples$pr, pw$samples$pr, 26),
            gap_mean(pm$samples$pr, pw$samples$pr, 56)))
```

## Recommended workflow

* **Diseases that compete** → `bamp_multicause()` (cause shares sum to all-cause; cross-cause
  trends coupled; emit additive per-person-year hazards for a microsimulation).
* **Exhaustive strata (sex, education, …)** → `bamp_coherent()` (shared trend + mean-reverting
  deviation; the between-stratum gap cannot diverge). `S > 2` strata are supported.
* **Both at once** → layer them: run `bamp_multicause()` within each sex (or stick-break causes on
  top of a sex-coherent total). See `docs/coherent-forecasting.md` for the nested cascade and the
  out-of-sample validation harness (`coherence_backtest()`, `select_rho()`).

The advantage over independent projection is not mainly short-horizon accuracy (the methods are
comparable there) but **coherence** — the parts add to the whole, the causes move together as the
data say, and the strata stay plausibly related — which is exactly what long-range, multi-disease,
multi-stratum projection needs.
```
